Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-05-01 Manuscript/manuscript.Rmd
edits to results section

We developed a computational framework<!--, refered to as *TADRF*,--> aimed at determining the optimal set of data level characteristics necessary for TAD boundary prediction using random forest classification models. We then leveraged our optimally built random forest to precisely predict the genomic coordinates of TAD boundaries at base pair resolution.

Our framework is resolution-specific and uses arbitrary called TAD boundaries as ground truth to establish TAD-boundary and non-TAD boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. TAD boundary regions were defined as regions on the linear genome where a genomic bin contained a called TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as genomic bins that did not contain a TAD boundary ($Y = 0$) (Figure 1A). Under this construction, the number of non-TAD boundary regions far exceed the number of TAD boundary regions. To cope with the imbalance among the classes of genomic bins, we evaluated three re-sampling techniques (in addition to no re-sampling) before predicting TAD boundaries. *Random over-sampling* (ROS) was defined as sampling with replacement from TAD boundary regions. *Random under-sampling* (RUS) was defined as sampling with replacement from non-TAD boundary regions. Lastly, we considered *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote].

A total of 78 cell type specific functional genomic annotations were used to build the feature space of downstream random forest models (Supplementary Table 2). The annotations could be broken down into three sets including histone modifications (HM), cis-regulatory elements (CRE), and transcription factor binding sites (TFBS). Four different feature engineering procedures were developed for each genomic element (Figure 1 B). *Peak Signal* was defined as the [???include definition]. Counts of overlaps (*Overlap Counts*, OC) were defined as the total number of genomic annotation regions that overlapped within a given genomic bin. Percents of overlaps (*Overlap Percent*, OP) were defined as the proportion of a bin overlapped by an annotation region. *Distance* was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). These feature engineering procedures were performed to improve TAD boundary predictions by accounting for direct (OC), proportional (Signal/OP), and spatial (distance) relationships between TAD boundaries and genomic annotations.

![**Illustrating the resolution-specific data construction and feature engineering used for downstream random forest modeling.** (A) The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a genomic bin that either overlapped with a TAD boundary ($Y = 1$) or did not ($Y = 0$). (B) Four types of associations between bins and genomic annotations (green shapes) were then used to build the feature space including _Peak Signal_ (Signal), _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_.](Figures/model_construction_and_predictor_types.png)

Thus, under our framework, we considered combinations of data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, with a specific engineered feature space *P = {Signal, OC, OP, Distance}*, and re-sampling technique *B = {None, RUS, ROS, SMOTE}* (Figure 2). Model inputs were evaluated by assessing the balanced accuracy (BA) of each combination. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $2 \times 5 \times 4 \times 4 (= 160)$ models. The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material.

![**A diagram of the ensemble framework detailing the steps for combining different inputs used in the random forest model building pipeline.** Step 1 is to establish the rowise (genomic bins) and columnwise (genomic annotations) dimensions of the data matrix, **A**, by combining a resolution (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)). Step 2 is to split **A** into a training and testing set according to a 7:3 ratio. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to implement the random forest algorithm and perform 3-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data and evaluate model performance using balanced accuracy (BA).](Figures/ensemble_framework.png)

We obtained a total of 8052, 7676, 4670, 2349, and 1031 called TADs across each resolution respectively (Table 1). From the called TADs, a total of 15468, 14253, 8363, 4224, and 1883 unique TAD boundaries were obtained (Table 2). The resolution-specific approach we developed to create TAD vs non-TAD boundary regions amenable for binary classification (See Methods) resulted in a total of 535363, 267682, 107073, 53537, and 26768 genomic bins. The class imbalance among the genomic bins was found to be both severe and invariant toward resolution. Specifically, we found the imbalance between TAD boundary and non-TAD boundary regions to be 0.03, 0.05, 0.08, 0.08, and 0.07 for each resolution respectively (Table 1). Furthermore, the data summaries presented in Table 1 were found to be similar to those exibited by the K562 cell line (Supplementary Table 3).

@@

<!--
Expectedly, when using data with class imbalance present, that is no re-sampling, the models exhibited low balanced accuracies, with very little variability among different resolutions (Figure 3). Similarly poor performances were found when using ROS as well. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS narrowly out performed SMOTE as the most optimal class balancing technique for all resolutions and predictor types. These results suggest that simple randomly under-sampling alone can drastically improve TAD boundary prediction. Additionally, we found that at higher resolutions, random forest classifiers built using a distance-type feature space yielded substantially higher MCCs when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with comparing class balancing techniques, this improvement was less evident at lower resolutions.

Interestingly, for the optimal re-sampling and predictor type, we found that performances peaked at 10 kb resolution and then degraded as data resolution decreased. The reason for the improved performance between 5 kb and 10 kb data resolutions can be attributed to the sparsity in the interaction frequency matrices at finer resolutions. This sparsity may lead to inaccurate TAD calling and, subsequently, to worse model performance. These results suggest that finer resolution of Hi-C data is not always optimal for accurate TAD calling and prediction, and sparsity of the data needs to be considered. Furthermore, performances tended to be much more variable between chromosomes at lower resolutions, evidenced by the wider error bars for the 50 kb and 100 kb resolutions. These results were found to be consistent for the K562 cell line (Supplementary Figure 2). In summary, our computational framework showed that using random under-sampling, distance-type predictors, and Hi-C data at 10kb resolution yields an optimally performing cell line-specific random forest classification model for TAD boundary prediction.
-->

![**Models built using distance-type predictors combined with random under-sampling at 5 kb yielded optimal predictive performance for GM12878.** Line plots comparing the performances of random forest classifiers built using different combinations of data level characteristics. Performances are compared across resolution, within each predictor-type: (A) OC, (B) OP, and (C) Distance, and between each re-sampling technique: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Performances were evaluated using balanced accuracy.](Figures/performances.png)

## Transcription factor binding sites outperform HM- and CRE-specific models

We then aimed to compare the performances of genomic class-specifc models. In doing so, we inspected the proportion of predictive importance that was explained by each of the three classes of genomic elements including histone modifications (HM), cis-regulatory elements (CRE), and transcription factor binding sites (TFBS). For an optimally built random forest classifier at each data resolution--smote re-sampling, and distance-type predictors--we found that the majority of the proportional predictive importance was obtained by transcription factor binding sites for both cell lines (Supplementary Figure 3 A-B). Furthermore, models built solely on TFBS outperformed other class-specific models (Figure 4A; Supplementary Figure 4A). Notably, the TFBS-specific models outperformed models built on the combination of all genomic elements at lower resolutions, indicating that the inclusion of other types of annotations contributed to additional noise rather than signal in TAD boundary prediction. Our results suggest that TFBS are the primary drivers of TAD boundary formation in both the GM12878 and K562 cell lines.

@@

Given that many TFBS are known to colocalize together on the linear genome and are therefore highly correlated, we sought to further optimize our TAD prediction model to avoid overfitting by performing recursive feature elimination. We found that, under our framework, we were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figure 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations were found to vary for each chromosome. To determine the most common TFBS, we clustered the predictive importances across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD boundary regions for all chromosomes (Figure 4C; Supplementary Figure 4C). Therefore, using our framework *TADRF* was able to identify the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@fudenberg2016formation; @alipour2012self]. Therefore, we could further optimize our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. To summarize, the optimally built cell line-specific random forest model under our *TADRF* framework was found to be one built using 10 kb resolution data, leveraging distance-type predictors for the CTCF, RAD21, SMC3, and ZNF143 transcription factor binding sites, while implementing random under-sampling to balance TAD and non-TAD boundary defined genomic bins.

![**Assessing top ranked transcription factors in TAD boundary prediction for GM12878.** (A) Barplots comparing performances of models using specific classes of functional genomic elements as the feature space including histone modifications (HM; red), cis-regulated elements (CRE; green), transcription factor binding sites (TFBS; blue), in addition to a model with a feature space containing all three classes (Combined; purple). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the mean MCC across each chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.](Figures/annots_specifc_rfe_importances.png)

## *preciseTAD* defines precise and biologically relevant locations of TAD boundaries

After obtaining our optimally built *TADRF* model, we investigated whether we could leverage our model to alleviate resolution limitations of Hi-C data by predicting TAD boundaries at base pair resolution. Our method, referred to as *preciseTAD*, uses a combination of complete linkage hierarchical clustering with a scalable version of partitioning around mediods (PAM), known as clustering large applications (CLARA) to reduce many adjacent base pairs with similarly high predictive probabilities down to a set of the most representative and biologically meaningful predicted TAD boundaries (see Methods). *preciseTAD* was applied to chromosome-specific sections of the linear genome. For each chromosome, we defined our new validation data using distance-type predictors among only the most informative TFBS for each base pair, namely the CTCF, RAD21, SMC3, and ZNF143 transcription factors. That is, each base pair had a corresponding row vector, **X**=$\{X_{1}, ..., X_{4}\}$, representing the $log_{2}$ genomic distances to the nearest respective transcription factor binding site.

We first used ROC analysis to obtain an optimal probability threshold for defining potential predicted TAD boundaries. We reported an average threshold of ??? (SD=???) among the 21 autosomal chromosomes that were analyzed (mean=???, SD=??? for K562). We referred to collections of chromosome-specific base pairs that exceeded this threshold when used to validate our *TADRF* model as *predicted TAD boundary areas* (PTBA), given that collections of adjacent base pairs tended to exhibit similar prediction profiles. We found that there were a total of ??? PTBA, containing an average of ??? base pairs each (??? PTBA for K562; mean base pairs=???, SD=???). *preciseTAD* then implements complete linkage hierarchical clustering on the dissimilarity matrix constructed from the pairwise distances (in base pairs) among the midpoints of each chromsome-specific PTBA. Using the silhouette score (see methods), we reported a total of k=??? chromosome-specific clusters (k=??? for K562). We referred to these *k* clusters as *predicted TAD boundary regions* (PTBR). That is, each PTBR contained the midpoints of multiple PTBA. Lastly, given each PTBR, *preciseTAD* uses CLARA partitioning to obtain the most representative midpoint. We referred to this as the *predicted TAD boundary point* (PTBP). These PTBP were identified as the most representative base pair among each PTBR. *preciseTAD* was able to identify a total of ??? PTBP across each chromosome. We reported less predicted TAD boundaries than ARROWHEAD at 10 kb (Table 1). This is attributed to ARROWHEAD's inherent inflation of called TAD boundaries, especially near fuzzy areas along the diagonal of Hi-C contact matrices (Figure 5A). Likewise, under the framework of *preciseTAD*, any two TADs are inherently demarcated by only one predicted TAD boundary, leading to comparatively less TAD boundaries. We also find that predicted TAD boundaries identified by *preciseTAD* were much more likely to demarcate high concentrations of inter-chromosomal contacts among Hi-C contact matrices (Figure 5A). [???include a result that demonstrates this; discuss in meeting]

@@

@@

For each predicted TAD boundary region, we were able to identify the summit locations of TAD boundary location by performing partitioning around medoids (see Methods). To compare the summit locations of predicted and called TAD boundaries, we evaluated the signal strength of ChIP-seq peak regions around a 10 kb flanking region. We found that the signal strength for the CTCF, RAD21, SMC3, and ZNF143 [??? Make all in this order!; !!!fixed] TFBS peaked at our predicted TAD boudaries, while was poorly defined around the called TAD boundaries (Figure 5B; Supplementary Figure 6).  Thus, our model is able to alleviate resolution limitations and precisely locate TAD boundaries that correspond to the highest signal strength of known TAD boundary-forming TFBSs. Furthermore, when compared to called TAD boundaries, the boundaries that we predicted with our model were found to be statistically significantly closer to the top ranked TFBS, suggesting stronger enrichment profiles near predicted TAD boundaries (Supplementary Figure 7).
-->

![**Predicted TAD boundaries are enriched in known TAD boundary-forming TFBSs.** (A) Visualizing locations of predicted and called TAD boundaries (chr 22: 17389000-18011000; GM12878). The upper triangular contact map shows the normalized number of contacts within the specified window using the JUICER toolbox.The blue vertical lines represent called boundaries identified by ARROWHEAD, with the corresponding called TADs outlined in blue on the upper diagonal contact matrix above. The black line represents the predicted probability of each base pair being a TAD boundary when validated by *TADRF*. Collections of red vertical lines represent base pairs whose predictive probability exceed the ROC threshold, labeled as *predicted TAD boundary areas* (PTBA). The midpoints of spatially similar PTBA were clustered together into *predicted TAD boundary regions* (PTBR), denoted by the yellow regions along the x-axis. The green vertical lines represent the *predicted TAD boundary points* (PTBA) identified by *preciseTAD*, with the corresponding predicted TADs outlined above in green.](Figures/visualizing_ptbp_and_ctbp.png)

@@

**Supplementary Table 1. Hi-C and genomic annotation data.** (A) A complete list of the Hi-C data libraries used to obtain the locations of topologically associating domains. (B) A complete list of the cell line specific genomic annotations used to build the feature space for all downstream models. The GRCh37/hg19 human genome assembly was used. "Genomic Class" - broad category of genomic features, "Element" - names of genomic annotations, "Cell Type-Specific Source" - download URL specific to the cell line (not all annotations were provided by the same institutions).

**Supplementary Figure 1. Performing $log_{2}$ transformations normalizes skewed distance densities across resolution.** When measured in basepairs, distances were highly right skewed (red lines) across all resolutions for both the GM12878 (A) and K562 (B) cell lines. After $log_{2}$ transforming the distances (blue lines), densities became much more gaussian distributed for all resolutions.

**Supplementary Figure 2. Models built using distance-type predictors combined with SMOTE re-sampling yielded optimal MCCs across resolution on K562.** Barplots comparing the performances of random forest classifiers built using different combinations of data level characteristics for K562. Within each re-sampling technique (A-D), performances are compared for each type of feature space: OC (red), OP (green), and Distance (blue), across resolutions. Performances were evaluated using MCCs and aggregated by averaging across each chromosome, with error bars representing 1 standard deviation away from the mean.

**Supplementary Figure 3. The proportional importance of optimally built cell line specific models was outweighed by transcription factor binding sites across resolutions.** Stacked barplots illustrate the proportion of total predictive importance in cell line specific random forest classifiers for (A) GM12878 and (B) K562 respectively. Model inputs included the optimal data level characteristics determined by the ensemble framework (SMOTE re-sampling; with distance-type predictors). Predictive importance was averaged for models built on chromosome specific regions of the linear genome. Proportional importance was then calculated according to the annotation group of the feature space including histone modifications (HM), chromatin state segmentations (ChromHMM), and transcription factor binding sites (TFBS).

**Supplementary Figure 4. Assessing top ranked transcription factors in TAD boundary prediction for K562.** (A) Barplots comparing performances of models using K562-specific groups of functional genomic annotation as the feature space including histone modifications (HM; red), chromatin states (ChromHMM; green), transcription factor binding sites (TFBS; blue), in addition to a model with a feature space containing all 3 groups (Combined; purple). Error bars represent ± 1 standard deviation from the mean MCC across each chromosomal section of the linear genome. (B) Recursive feature elimination analysis used to reduce the number of TFBS to include in the optimally build random forest classifier. Error bars represent 1 standard deviation away from the mean cross-validated MCC. (C) Clustered heatmap of predictive importance for the union of the top 8 most predictive TFBS across chromosomes.

>>>>

We developed a computational framework<!--, refered to as *TADRF*,--> aimed at determining the optimal set of data level characteristics necessary for topologically associated domain (TAD) boundary prediction using random forest classification models. We then leveraged our optimally built random forest to precisely predict the genomic coordinates of both TAD and loop boundaries at base pair resolution.

Our framework is resolution-specific and uses arbitrary called TAD and loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries we opted to use boundaries obtained by the PEAKACHU tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised].

Boundary regions were defined as regions on the linear genome where a genomic bin contained a called boundary ($Y = 1$), while non-boundary regions were defined as genomic bins that did not contain a boundary ($Y = 0$) (Figure 1A). Under this construction, the number of non-boundary regions far exceeded the number of boundary regions. To cope with the imbalance among the classes of genomic bins, we evaluated three re-sampling techniques (in addition to no re-sampling) before predicting boundary regions. *Random over-sampling* (ROS) was defined as sampling with replacement from boundary regions. *Random under-sampling* (RUS) was defined as sampling with replacement from non-boundary regions. Lastly, we considered *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote].

A total of 78 cell line specific functional genomic annotations were used to build the feature space of downstream random forest models (Supplementary Table 2). The annotations could be broken down into three sets including histone modifications (HM), cis-regulatory elements (BroadHMM), and transcription factor binding sites (TFBS). Four different feature engineering procedures were developed for each genomic element (Figure 1 B). *Peak Signal* was defined as the [???include definition]. Counts of overlaps (*Overlap Counts*, OC) were defined as the total number of genomic annotation regions that overlapped within a given genomic bin. Percents of overlaps (*Overlap Percent*, OP) were defined as the proportion of a bin overlapped by an annotation region. *Distance* was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). These feature engineering procedures were performed to improve boundary predictions by accounting for direct (OC), proportional (Signal/OP), and spatial (distance) relationships between TAD boundaries and genomic annotations.

![**Illustrating the resolution-specific data construction and feature engineering used for downstream random forest modeling.** (A) The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a genomic bin that either overlapped with a TAD boundary ($Y = 1$) or did not ($Y = 0$). (B) Four types of associations between bins and genomic annotations (green shapes) were then used to build the feature space including _Peak Signal_ (Signal), _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_.](Figures/figure1_model_construction_and_predictor_types.png)

Thus, under our framework, we considered combinations of data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, with a specific engineered feature space *P = {Signal, OC, OP, Distance}*, and re-sampling technique *S = {None, RUS, ROS, SMOTE}* (Figure 2). Once the model inputs were established, a random forest classifier was trained on chromosome-specific genomic bins (excluding chromosome 9), *C = {CHR1, ..., CHR22}*, using 5-fold cross-validation. Model inputs were evaluated by assessing the balanced accuracy (BA) of each combination. <!--The results presented are for the ARROWHEAD boundaries, with PEAKACHU boundaries presented in the Supplementary material. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |S| \times |C|$ different ensemble systems, totaling $2 \times 5 \times 4 \times 4 \times 21 (= 3360)$ models.-->

![**A diagram of the ensemble framework detailing the steps for combining different inputs used in the random forest model building pipeline.** Step 1 is to establish the chromosome-specific rowise (genomic bins) and columnwise (genomic annotations) dimensions of the data matrix, **A**, by combining a resolution (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)). Step 2 is to split **A** into a training and testing set according to a 7:3 ratio. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to implement the random forest algorithm and perform 5-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data and evaluate model performance using balanced accuracy (BA).](Figures/figure2_ensemble_framework.png)

Using ARROWHEAD, we obtained a total of 8052, 7676, 4670, 2349, and 1031 called TADs  across each resolution respectively (Table 1). From the called TADs, a total of 15468, 14253, 8363, 4224, and 1883 unique TAD boundaries were obtained (Table 1). The resolution-specific approach we developed to create boundary vs non-boundary regions resulted in a total of 535363, 267682, 107073, 53537, and 26768 genomic bins. The class imbalance among the genomic bins was both severe and invariant toward resolution, resulting in 0.03, 0.05, 0.08, 0.08, and 0.07 for each resolution respectively (Table 1). Boundary data summaries were similar for K562 cell line, as well as for PEAKACHU defined boundaries (Supplementary Tables 3-4).

@@

Expectedly, when using data with class imbalance present, that is no re-sampling, the models exhibited low balanced accuracies, with very little variability among different resolutions (Figure 3). Similarly poor performances were found when using ROS as well. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS narrowly out performed SMOTE as the most optimal class balancing technique for all resolutions and predictor types. These results suggest that simple randomly under-sampling alone can drastically improve TAD boundary prediction. Additionally, we found that at higher resolutions, random forest classifiers built using a distance-type feature space yielded substantially higher MCCs when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with comparing class balancing techniques, this improvement was less evident at lower resolutions. These results were found to be consistent for the K562 cell line (Supplementary Figure 2). Contrary to ARROWHEAD defined boundaries, random forest models built on PEAKACHU loop boundaries exhibited optimal performance when implemented on SMOTE re-sampled data for both GM12878 and K562 cell lines (Supplementary Figure 3). In summary, our computational framework showed that using random under-sampling, distance-type predictors, and Hi-C data at 5kb resolution yields an optimally performing cell line-specific random forest classification model for TAD boundary region prediction.

<!--
Interestingly, for the optimal re-sampling and predictor type, we found that performances peaked at 10 kb resolution and then degraded as data resolution decreased. The reason for the improved performance between 5 kb and 10 kb data resolutions can be attributed to the sparsity in the interaction frequency matrices at finer resolutions. This sparsity may lead to inaccurate TAD calling and, subsequently, to worse model performance. These results suggest that finer resolution of Hi-C data is not always optimal for accurate TAD calling and prediction, and sparsity of the data needs to be considered. Furthermore, performances tended to be much more variable between chromosomes at lower resolutions, evidenced by the wider error bars for the 50 kb and 100 kb resolutions. These results were found to be consistent for the K562 cell line (Supplementary Figure 2). In summary, our computational framework showed that using random under-sampling, distance-type predictors, and Hi-C data at 10kb resolution yields an optimally performing cell line-specific random forest classification model for TAD boundary prediction.
-->

![**Models built using distance-type predictors combined with random under-sampling at 5 kb yielded optimal predictive performance for GM12878.** Line plots comparing the performances of random forest classifiers built using different combinations of data level characteristics. Performances are compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and between each re-sampling technique: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Performances were evaluated using balanced accuracy.](Figures/figure3_performances.png)

## Transcription factor binding sites outperform HM- and BroadHMM-specific models

We then aimed to compare the performances of genomic class-specifc models. In doing so, we evaluated performances for an optimally built random forest model using only histone modifications (HM), cis-regulatory elements (BroadHMM), and transcription factor binding sites (TFBS) separately. We found that models built solely on TFBS outperformed other class-specific models suggesting that TFBS are the primary drivers of TAD and loop boundary formation in both the GM12878 and K562 cell lines (Figure 4A; Supplementary Figures 4A,5A,6A).

@@

We sought to further optimize our TAD prediction model by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all chromosomes. Under our framework, we were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B,5B,6B). However, given that we trained our model on chromosome-specific data, the most significant annotations were found to vary for each chromosome. To determine the most common TFBS, we clustered the predictive importances across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD boundary regions for all chromosomes (Figure 4C; Supplementary Figures 4C,5C,6C). Therefore, our framework was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@fudenberg2016formation; @alipour2012self]. We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance.

<!--
To summarize, the optimally built cell line-specific random forest model under our computational framework was found to be one built using 10 kb resolution data, leveraging distance-type predictors for the CTCF, RAD21, SMC3, and ZNF143 transcription factor binding sites, while implementing random under-sampling to balance TAD and non-TAD boundary defined genomic bins.
-->

![**Assessing top ranked transcription factors in TAD boundary prediction for GM12878.** (A) Barplots comparing performances of models using specific classes of functional genomic elements as the feature space including histone modifications (HM; red), cis-regulated elements (BroadHMM; green), transcription factor binding sites (TFBS; blue), in addition to a model containing all three classes (ALL; purple). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.](Figures/figure4_annots_specifc_rfe_importances.png)

## *preciseTAD* identifies precise and biologically relevant locations of TAD boundaries

After obtaining our optimally built random forest model, we investigated whether we could leverage our model to alleviate resolution limitations of Hi-C data by predicting TAD boundaries at base pair resolution. Our method, referred to as *preciseTAD*, leverages a user specified binary classifier with a combination of density-based hierarchical clustering and a scalable version of partitioning around mediods (PAM), known as clustering large applications (CLARA) to precisely identify the most representative and biologically meaningful predicted chromosome-specific domain boundaries (see Methods).

Here, we opted to implement preciseTAD using an optimally built whole genome random forest model. That is, a random forest model built on 5 kb defined genomic bins, with random under-sampling, and distance-type predictors for the top informative TFBS (CTCF, RAD21, SMC3, and ZNF143). The reason for this was because of the improved performance that was exhibited for a whole genome random forest compared to aggregated chromosome-specific models (Supplementary Figure 7).

@@

<!--
We first used ROC analysis to obtain an optimal probability threshold for defining potential predicted TAD boundaries. We reported an average threshold of ??? (SD=???) among the 21 autosomal chromosomes that were analyzed (mean=???, SD=??? for K562). We referred to collections of chromosome-specific base pairs that exceeded this threshold when used to validate our *TADRF* model as *predicted TAD boundary areas* (PTBA), given that collections of adjacent base pairs tended to exhibit similar prediction profiles. We found that there were a total of ??? PTBA, containing an average of ??? base pairs each (??? PTBA for K562; mean base pairs=???, SD=???). *preciseTAD* then implements complete linkage hierarchical clustering on the dissimilarity matrix constructed from the pairwise distances (in base pairs) among the midpoints of each chromsome-specific PTBA. Using the silhouette score (see methods), we reported a total of k=??? chromosome-specific clusters (k=??? for K562). We referred to these *k* clusters as *predicted TAD boundary regions* (PTBR). That is, each PTBR contained the midpoints of multiple PTBA. Lastly, given each PTBR, *preciseTAD* uses CLARA partitioning to obtain the most representative midpoint. We referred to this as the *predicted TAD boundary point* (PTBP). These PTBP were identified as the most representative base pair among each PTBR. *preciseTAD* was able to identify a total of ??? PTBP across each chromosome. We reported less predicted TAD boundaries than ARROWHEAD at 10 kb (Table 1). This is attributed to ARROWHEAD's inherent inflation of called TAD boundaries, especially near fuzzy areas along the diagonal of Hi-C contact matrices (Figure 5A). Likewise, under the framework of *preciseTAD*, any two TADs are inherently demarcated by only one predicted TAD boundary, leading to comparatively less TAD boundaries. We also find that predicted TAD boundaries identified by *preciseTAD* were much more likely to demarcate high concentrations of inter-chromosomal contacts among Hi-C contact matrices (Figure 5A). [???include a result that demonstrates this; discuss in meeting]
-->

@@

For each predicted TAD boundary region, we were able to identify the summit locations of TAD boundary location by performing partitioning around medoids (see Methods). To compare the summit locations of predicted and called TAD boundaries, we evaluated the signal strength of ChIP-seq peak regions around a 10 kb flanking region. We found that the signal strength for the CTCF, RAD21, SMC3, and ZNF143 [??? Make all in this order!; !!!fixed] TFBS peaked at our predicted TAD boudaries, while was poorly defined around the called TAD boundaries (Figure 5B; Supplementary Figure 6).  Thus, our model is able to alleviate resolution limitations and precisely locate TAD boundaries that correspond to the highest signal strength of known TAD boundary-forming TFBSs. Furthermore, when compared to called TAD boundaries, the boundaries that we predicted with our model were found to be statistically significantly closer to the top ranked TFBS, suggesting stronger enrichment profiles near predicted TAD boundaries (Supplementary Figure 7).

![**Predicted TAD boundaries are enriched in known TAD boundary-forming TFBSs.** (A) Visualizing locations of predicted and called TAD boundaries (chr 22: 17389000-18011000; GM12878). The upper triangular contact map shows the normalized number of contacts within the specified window using the JUICER toolbox.The blue vertical lines represent called boundaries identified by ARROWHEAD, with the corresponding called TADs outlined in blue on the upper diagonal contact matrix above. The black line represents the predicted probability of each base pair being a TAD boundary when validated by *TADRF*. Collections of red vertical lines represent base pairs whose predictive probability exceed the ROC threshold, labeled as *predicted TAD boundary areas* (PTBA). The midpoints of spatially similar PTBA were clustered together into *predicted TAD boundary regions* (PTBR), denoted by the yellow regions along the x-axis. The green vertical lines represent the *predicted TAD boundary points* (PTBA) identified by *preciseTAD*, with the corresponding predicted TADs outlined above in green.](Figures/visualizing_ptbp_and_ctbp.png) 
-->

@@

**Supplementary Table 1.** A complete list of the Hi-C data libraries obtained from GEO GSE63525 [@rao20143d] used to obtain the locations of topologically associating domains for ARROWHEAD, as well as the locations for the cell line specific .txt files of loop boundaries obtained by PEAKACHU [@salameh2019supervised].

**Supplementary Table 2.** A complete list of the genomic annotations used to build the feature space for all downstream models. The GRCh37/hg19 human genome assembly was used. "Genomic Class" - broad category of genomic features, "Element" - names of genomic features, "Cell Type-Specific Source" - download URL specific to the cell line (not all annotations were provided by the same institutions).

**Supplementary Figure 1.** Density curves illustrate the normalization of distances before (red) and after (blue) performing a $log_{2}$ transformation across differing bin sizes including 10 kb, 25 kb, 50 kb, and 100 kb for both the (A) GM12878 and (B) K562 cell lines respectively. Distances are measured as the number of base pairs from the center of a genomic bin to the center of the nearest genomic annotation. Each density curve represents a particular genomic annotation.

**Supplementary Table 3.** Domain boundary and class imbalance summaries from SpectralTAD across resolution for both the
GM12878 and K562 cell lines respectively.

**Supplementary Table 4.** Loop boundary and class imbalance summaries from PEAKACHU 10 kb resolution for both the GM12878 and K562 cell lines respectively.

**Supplementary Figure 2. Models built using distance-type predictors combined with random under-sampling at 5 kb yielded optimal predictive performance for K562.** Line plots comparing the performances of random forest classifiers built using ARROWHEAD boundaries at different combinations of data level characteristics. Performances are compared across resolution, within each predictor-type: (A) Signal (B) OC, (C) OP, and (D) Distance, and between each re-sampling technique: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Performances were evaluated using balanced accuracy.

**Supplementary Figure 3. Models built using distance-type predictors combined with SMOTE re-sampling yielded optimal predictive performance for GM12878 K562.** Line plots comparing the performances of random forest classifiers built using using PEAKACHU boundaries at different combinations of data level characteristics for (A) GM12878 and (B) K562. Performances are compared across resampling technique: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple), and within each predictor-type: Signal, OC, OP, and Distance.

**Supplementary Figure 4. Assessing top ranked transcription factors in TAD boundary prediction for K562.** (A) Barplots comparing performances of models built using ARROWHEAD boundaries on specific classes of functional genomic elements as the feature space including histone modifications (HM; red), cis-regulated elements (BroadHMM; green), transcription factor binding sites (TFBS; blue), in addition to a model containing all three classes (ALL; purple). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.
